{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-05-02 paper.tex
first version of the conclusions

\begin{abstract}
We present a new software package designed to aid the inclusion of hadronic
observables into Parton Distribution Function (PDF) fits. The {\tt APFELgrid} package
converts interpolated weight tables provided by {\tt APPLgrid} files into a more efficient
format for PDF fitting by the pre-computation and combination with PDF and $\alpha_s$ evolution factors provided
by {\tt APFEL}. This combination significantly reduces the number of operations required to perform the calculation
of hadronic observables in PDF fits and simplifies the structure of the calculation into
a readily optimised scalar product. We demonstrate that our technique
can lead to a substantial speed improvement as compared to the
existing interfaces without any reduction in numerical accuracy.

%We present a new method conceived to ease the inclusion of hadronic
%observables into parton distribution function (PDF) fits. Our
%technique relies on the the same principles adopted by the existing
%fast interfaces but implements a set of improvements aimed at
%optimizing the performance in the context of PDF fits where many
%iterations and many predictions for each iteration are usually
%required. The main novelties are the precomputation of the PDF and
%$\alpha_s$ evolution and the optimisation of the numerical convolution
%between hard cross sections and PDFs. We demonstrate that our technique
%can lead to a substantial speed improvement as compared to the
%existing interfaces.
\end{abstract}

@@

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the hard
cross-sections, these packages rely on external codes to which they
are interfaced by means of a suite of
functions allowing for the filling of PDF- and
$\alpha_s$-independent look-up tables of cross-section weights. Codes
such as {\tt MCFM}~\cite{Campbell:2010ff} and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to 
 {\tt APPLgrid}/{\tt FastNLO} and more recently dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. The {\tt aMCfast} and {\tt MCgrid} codes
have been specifically designed to extract the relevant information
from the {\tt MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

@@

\subsection{The FastKernel method}
A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa,Carrazza:2014gfa,Bertone:2015cwa,Bertone:2015lqa,Carrazza:2015dea,Bertone:2015gba}
or {\tt HOPPET}~\cite{Salam:2008qg}, are available which perform PDF
evolution via an analogous procedure. {\tt APFEL} in particular, is
able to provide the so-called \textit{evolution operator}. The
evolution operator, is computed assuming that the DGLAP evolution can
be encoded in a single operator whose action on the intial scale PDFs
proceduces the final scale PDFs. Taking $Q_0$ as an initial scale and
$Q_\tau$ as a final scale, on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution
equation with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

@@

\section{Conclusion}
\label{sec:conclusion}

In this work we presented {\tt APFELgrid}, an efficient convolution
interface to hadronic observables for fits of parton densities.

>>>>

\begin{abstract}
  We present a new software package designed to aid the inclusion of
  hadronic observables into Parton Distribution Function (PDF)
  fits. The {\tt APFELgrid} package converts interpolated weight
  tables provided by {\tt APPLgrid} files into a more efficient format
  for PDF fitting by the pre-computation and combination with PDF and
  $\alpha_s$ evolution factors provided by {\tt APFEL}. This
  combination significantly reduces the number of operations required
  to perform the calculation of hadronic observables in PDF fits and
  simplifies the structure of the calculation into a readily optimised
  scalar product. We demonstrate that our technique can lead to a
  substantial speed improvement as compared to the existing interfaces
  without any reduction in numerical accuracy.
  % We present a new method conceived to ease the inclusion of
  % hadronic observables into parton distribution function (PDF)
  % fits. Our technique relies on the the same principles adopted by
  % the existing fast interfaces but implements a set of improvements
  % aimed at optimizing the performance in the context of PDF fits
  % where many iterations and many predictions for each iteration are
  % usually required. The main novelties are the precomputation of the
  % PDF and $\alpha_s$ evolution and the optimisation of the numerical
  % convolution between hard cross sections and PDFs. We demonstrate
  % that our technique can lead to a substantial speed improvement as
  % compared to the existing interfaces.
\end{abstract}

@@

This interpolation strategy is implemented in the {\tt
  APPLgrid}~\cite{Carli:2010rw} and {\tt
  FastNLO}~\cite{Wobisch:2011ij} projects. For the computation of the
hard cross-sections, these packages rely on external codes to which
they are interfaced by means of a suite of functions allowing for the
filling of PDF- and $\alpha_s$-independent look-up tables of
cross-section weights. Codes such as {\tt MCFM}~\cite{Campbell:2010ff}
and {\tt NLOJet++}~\cite{Nagy:2003tz} have been interfaced directly to
{\tt APPLgrid}/{\tt FastNLO} and more recently dedicated interfaces to
automated general-purpose Monte Carlo (MC) event generators have been
developed. The {\tt aMCfast}~\cite{Bertone:2014zva} and {\tt
  MCgrid}~\cite{DelDebbio:2013kxa} codes have been specifically
designed to extract the relevant information from the {\tt
  MadGraph5\_aMC@NLO}~\cite{Alwall:2014hca} and {\tt
  SHERPA}~\cite{Gleisberg:2008ta} event generators respectively.

@@

\subsection{The FastKernel method}

A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa,Carrazza:2014gfa,Bertone:2015cwa,Bertone:2015lqa,Carrazza:2015dea,Bertone:2015gba}
or {\tt HOPPET}~\cite{Salam:2008qg}, are available which perform PDF
evolution via an analogous procedure. {\tt APFEL} in particular, is
able to provide the so-called \textit{evolution operator}. The
evolution operator, is computed assuming that the DGLAP evolution can
be encoded in a single operator whose action on the intial scale PDFs
proceduces the final scale PDFs. Taking $Q_0$ as an initial scale and
$Q_\tau$ as a final scale, on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution equation
with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

@@

\section{Conclusion}\label{sec:conclusion}

In this work we have presented {\tt APFELgrid}, an efficient
convolution interface to hadronic observables for fits of parton
densities. We have demonstrated that employing the so-called {\tt
  FastKernel} method, it is possible to convert an {\tt APPLgrid}
weight table into a derived format, referred to as {\tt FKtable},
which includes PDF and $\alpha_s$ evolution as provided by {\tt
  APFEL}.  Such a procedure allows one to obtain a very
computationally-efficient expression for the calculation of hadronic
cross sections, in terms of only the initial-scale PDFs and summing
only over the initial scale interpolating grid and the incoming parton
flavors. In addition, the simple structure of the resulting
convolution makes {\tt FKtables} particularly suitable to make
efficient use of acceleration and parallelization techniques like {\tt
  SIMD} and {\tt OpenMD}.

We have shown that the numerical convolution of an {\tt APFELgrid}
tables is typically much faster than the corresponding {\tt
  APPLgrid}. In particular, the {\tt FastKernel} method provides a
great perfomance improvement with respect to {\tt APPLgrid} which is
further enhanced by the employment of any of the acceleration
techniques available in {\tt APFELgrid}. Finally, while the size of an
{\tt FKtable} is often larger than the corresponding {\tt APFELgrid}
file, we have shown that the resident set size memory required by {\tt
  APFELgrid} is typically smaller than {\tt APPLgrid} by at least
75\%.

The subtantial speed improvement with respect to {\tt APPLgrid} in
association with the reduction of the resident set size memory, makes
the {\tt APFELgrid} code a valuable tool for modern PDF fits which
usually incluse thousands of hadronic data points and that thus
heavily rely on fast interfaces. We stress that {\tt APFELgrid} is not
a replacement for {\tt APPLgrid}, on which it depends, but it is
rather a specialized tool conceived to ease the inclusion of hadronic
data into PDF fits. In fact, the price of a better performance is the
fact that, by themselves, {\tt FKtables} cannot perform any parameter
variation other than that of PDFs. For example, variations of
$\alpha_s$ can be performed only by re-computing the {\tt FKtables}.

The {\tt APFELgrid} code is publicly available from the webpage:
\begin{center}
\url{https://github.com/nhartland/APFELgrid}
\end{center}
where intructions to install and run codes are also provided.
